A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2023-12-27 paper/paper.tex
Add preservation (WIP)

@@

The transitivity of the coherence relation follows as a corollary of
Lemma~\ref{lemma:parconfluent}.
\begin{lemma}[Coherence trans]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
By the definition of coherence, there exists some term $[[b]]$ such that $[[a0 =>+ b0]]$,
$[[a1 =>+ b0]]$ and some term $[[b1]]$ such that $[[a1 =>+ b1]]$ and
$[[a2 =>+ b1]]$. By Lemma~\ref{lemma:parconfluent}, there exists some
term $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$. It sufficies
to show that $[[a0 =>+ c]]$ and $[[a2 =>+ c]]$, both of which
trivially hold since the transitive closure $[[a =>+ b]]$ is transitive.
This concludes the proof that coherence is an equivalence relation.

@@

The relation $[[Interp I i A S]]$ satisfies the following inversion
principles.
\begin{lemma}[Void Inv]
  \label{lemma:voidinv}
  If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
\end{lemma}
\begin{lemma}[Bool Inv]
  \label{lemma:boolinv}
  If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
\end{lemma}
\begin{lemma}[Eq Inv]
  \label{lemma:eqinv}
  If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
\end{lemma}
\begin{lemma}[Pi Inv]
  \label{lemma:piinv}
  If $[[Interp I i Pi A B S]]$, then there exists $[[S]],[[F]]$ such
  that the following conditions hold:
  \begin{itemize}
  \item $[[Interp I i A S ]]$
  \item $[[F in S -> PowerSet STm]]$
  \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
  \end{itemize}
\end{lemma}
\begin{lemma}
  \label{lemma:setinv}
  If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
\end{lemma}

>>>>

From Lemma~\ref{lemma:parcong} and \ref{lemma:parrefl}, we recover the
single substitution property as a simple corollary.
\begin{corollary}[Par subst]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
\end{corollary}

@@

The transitivity of the coherence relation follows as a corollary of
Lemma~\ref{lemma:parconfluent}.
\begin{lemma}[Coherence trans]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
By the definition of coherence, there exists some term $[[b]]$ such that $[[a0 =>+ b0]]$,
$[[a1 =>+ b0]]$ and some term $[[b1]]$ such that $[[a1 =>+ b1]]$ and
$[[a2 =>+ b1]]$. By Lemma~\ref{lemma:parconfluent}, there exists some
term $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$. It sufficies
to show that $[[a0 =>+ c]]$ and $[[a2 =>+ c]]$, both of which
trivially hold since the transitive closure $[[a =>+ b]]$ is transitive.
This concludes the proof that coherence is an equivalence relation.
\begin{lemma}[Coherence Equivalence]
  \label{lemma:coherenceequiv}
  The relation $[[a <=> b]]$ satisfies reflexivity, symmetry, and
  transitivity and therefore is an equivalence relation.
\end{lemma}

@@

The relation $[[Interp I i A S]]$ satisfies the following inversion
principles.
\begin{lemma}[Inversion of the logical relation]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
  \item If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
  \item If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
  \item If $[[Interp I i Pi A B S]]$, then there exists $[[S]],[[F]]$ such that:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
  \item If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  We only show the most involved function case; the rest follows a
  similar but simpler pattern. We start by
  inducting over the derivation of $[[Interp I i Pi A B S]]$. There
  are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:] We know that there exists some $[[A0]]$ and
    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$, and from the
    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
    \end{itemize}
    By inverting $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a} => B0 {a} ]]$ for all
    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
    $[[F]]$ satisfies the desired properties:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
  \end{description}
\end{proof}

\Rref{I-Red} bakes into the logical relation the backward preservation
property. The following property shows that preservation holds in the
usual forward direction too.
\begin{lemma}[Preservation of the logical relation]
  \label{lemma:interppreservation}
  If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
\end{lemma}
\begin{proof}
  We carry out the proof by induction over the derivation of $[[Interp
  I i A S]]$.
  \begin{description}
  \item[\Rref{I-Void}:] There exists some $[[B]]$ such that $[[Void =>
    B]]$. By inverting the derivation of $[[Void => B]]$, $[[B]]$ must
    be $[[Void]]$ and the result trivially follows.
  \item[\Rref{I-Bool, I-Set}:] Similar to the case for \rref{I-Void}.
  \item[\Rref{I-Eq}:] We know that $[[Interp I i a ~ b : A { p | p =>+
      refl , a <=> b  }]]$ and, by inverting the derivation of
    parallel reduction, $[[a => a0]]$, $[[b => b0]]$, $[[A => A0]]$
    for some $[[a0]]$, $[[b0]]$, and $[[A0]]$. Our goal is to show
    that $[[Interp I i a0 ~ b0 : A {p | p =>+ refl, a <=> b}]]$. By
    \rref{I-Eq}, we already know that $[[Interp I i a0 ~ b0 : A {p | p
      =>+ refl, a0 <=> b0}]]$ and therefore it suffices to show that
    the sets $[[{p | p =>+ refl, a <=> b}]]$ and $[[{p | p =>+ refl,
      a0 <=> b0}]]$ are equal. Equivalently, it suffices to show that
    $[[a <=> b]]$ if and only if $[[a0 <=> b0]]$. By definition, from
    $[[a => a0]]$ and $[[b => b0]]$, we derive $[[a <=> a0]]$ and $[[b
    <=> b0]]$. The result then immediately follows from the fact that
    coherence is an equivalence relation
    (Lemma~\ref{lemma:coherenceequiv}).
  \item[\Rref{I-Pi}:]
  \end{description}
\end{proof}
